\huge{ \emph{Supplement to} Algorithms For Shaping a Particle Swarm\\ With a Shared Control Input Using Boundary Interaction}

### 2016-11-18 RA-L/theory.tex
Added Changes

\subsection{Using Boundaries: stable configurations of a swarm}\label{subsec:FluidInTank}
One method to control a swarm's shape in a bounded workspace is to simply push in a given direction until the swarm conforms to the boundary.
\paragraph{Square workplace}
%The formulas for means $(\bar{x},\bar{y})$, covariance $(\sigma^2_x,\sigma^2_y,\sigma_{xy})$, and correlation $\rho_{xy}$ are as follows: 
We first examine the mean $(\bar{x},\bar{y})$, covariance $(\sigma^2_x,\sigma^2_y,\sigma_{xy})$, and correlation $\rho_{xy}$ of a very large swarm of robots as they move inside a square workplace under the influence of gravity pointing in the direction $\beta$. The swarm is large, but the robots are small in comparison, and together occupy a constant area $A$ ranging from [0,1]. Under a global input such as gravity, they flow like water, moving to a side of the workplace and forming a polygonal shape, as shown in Fig.~\ref{fig:SquareFill},  see also \citep{Zhao2016mathematicaSquare}.

@@

The region of integration $R$ is the polygon containing the swarm. If $A<1/2$ and the force angle is $\beta$, the mean when $R$ is a triangular region in the lower-left corner is:
\begin{align}\label{eq:meanInSquareWorkspaceLL}
\bar{x}(A,\beta) &= \frac{\int_0^{\sqrt{2 A \tan (\beta)}} \left(\int_0^{\cot (\beta) \left(\sqrt{2 A \tan (\beta)}-x\right)} x \, dy\right) \, dx}{A} \nonumber \\
	&=\frac{ \sqrt{2}}{3} \sqrt{A \tan (\beta )}\\
\bar{y}(A,\beta) &= \frac{\int_0^{\sqrt{2 A \tan (\beta)}} \left(\int_0^{\cot (\beta) \left(\sqrt{2 A \tan (\beta)}-x\right)} y \, dy\right) \, dx}{A} \nonumber\\
	&=\frac{\sqrt{2}}{3}  \sqrt{A \cot (\beta )}
\end{align}
The full equations are included in the appendix~\citep{2016arXiv160901830S}, and are summarized in Fig.~\ref{fig:SquareFill}. A few highlights are that the correlation is maximized $\pm$1/2 when the swarm is in a triangular shape. The covariance of a triangle is always $\pm(A/18)$. Variance is minimized in the direction of $\beta$ and maximized orthogonal to $\beta$ when the swarm is in a rectangular shape. The range of mean positions are maximized when $A$ is small.

@@

Let the control input be a vector force $\vec{F}$ with magnitude $F$ and orientation $\theta$ with respect to a line perpendicular to and into the nearest boundary. $N$ is the normal or perpendicular force between the robot and the boundary. The force of friction $F_f$ is nonzero if the robot is in contact with the boundary and  $|\theta| < \pi/2$. The resulting net force on the robot, $F_{\text{\emph{forward}}}$, is aligned with the wall and given by
\begin{align}
F_{\text{\emph{forward}}} &=  F \sin(\theta) - F_f  \nonumber \\
\text{where }  F_f &= \begin{cases}  \mu_f N, &  \mu_f N < F \sin(\theta)  \label{eq:frictionmodel}  \\
F \sin(\theta), & \text{else} \end{cases} \\ %\sign(F \sin(\theta) ) \cdot  \max(0, | F sin\theta |- |F_f|)
\text{and } N &= F \cos(\theta) \nonumber
\end{align}
 Fig.~\ref{fig:friction} shows the resultant forces on two robots when one is touching a wall. 
Though each receives the same inputs,  they experience different net forces.
  For ease of analysis, the following algorithms assume $\mu_f$ is infinite and robots touching the wall are prevented from sliding along the wall.
This means that if one robot is touching the wall and another robot is free, the touching robot will not move when the control input is parallel or into the wall. 
There are many alternate models of friction that also break control symmetry. Fig.~\ref{fig:friction}c shows fluid flow along a boundary.  Fluid in the free-flow region moves uniformly, but flow decreases to zero in the boundary layer.  
\begin{align}
%u(y) = u_0 \left(1- \frac{(y-h)^2}{h^2}\right) = u_0 \frac{y}{h} \left(2- \frac{y}{h}\right) \label{eq:boundarylayerflow} THiS EQUATION IS TERRIBLE! PLOT IT.
F_{forward}(y) &= F - F_f\begin{cases}  \left( \frac{y-h}{h} \right)^2  , &  y<h \label{eq:boundarylayerflow} \\
0, & \text{else} \end{cases}
\end{align}

@@

\section{Algorithms}\label{sec:algorithms}
\subsection{Position Control of $2$ Robots Using Wall Friction}\label{sec:PostionControl2Robots}
\begin{figure*}
\centering
\renewcommand{\figwid}{0.4\columnwidth}
{\begin{overpic}[width =\figwid]{S1_1.png}\put(10,15){$t$ = 0 s}
\end{overpic}
\begin{overpic}[width =\figwid]{S1_2.png}\put(10,15){$t$ = 0.14 s}
\end{overpic}
\begin{overpic}[width =\figwid]{S1_3.png}\put(10,15){$t$  = 0.29 s}
\end{overpic}
\begin{overpic}[width =\figwid]{S1_4.png}\put(10,15){$t$  = 0.43 s}
\end{overpic}
\begin{overpic}[width =\figwid]{S1_5.png}\put(10,15){$t$  = 0.76 s}
\end{overpic}}\\
%{\begin{overpic}[width =\figwid]{two_1.jpg}\put(45,75){$t$  = 0 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{two_2.jpg}\put(45,75){$t$  = 0.25 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{two_3.jpg}\put(45,75){$t$  = 0.5 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{two_4.jpg}\put(45,75){$t$  = 0.75 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{two_5.jpg}\put(45,75){$t$  = 1 s}
%\end{overpic}}\\
%\vspace{.5em}
%{\begin{overpic}[width =\figwid]{S3_1.jpg}\put(45,75){$t$ = 0 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{S3_2.jpg}\put(45,75){$t$ = 0.25 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{S3_3.jpg}\put(45,75){$t$  = 0.5 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{S3_4.jpg}\put(45,75){$t$  = 0.75 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{S3_5.jpg}\put(45,75){$t$  = 1 s}
%\end{overpic}}
%\vspace{-1em}
\caption{\label{fig:shapeControlMathematica1}{Frames from an implementation of Alg.\ \ref{alg:PosControl2Robots}: two robot positioning using walls with infinite friction. The algorithm only requires friction along the bottom and right walls.
Robot initial positions are shown by a crosshair, and final positions by a circled crosshair.  Dashed lines show the shortest route if robots could be controlled independently.  Solid arrows show path given by  Alg.\ \ref{alg:PosControl2Robots}.
Online demonstration and source code at \citep{Shahrokhi2015mathematicaParticle}.
%The bottom row shows an extreme case where the robots must switch position.
}
%\vspace{-2em}
}
\end{figure*}

@@

%Step (i): Fixing $\Delta r_x$
%\label{theory:step1}
%\begin{enumerate}
%\item
%If $\Delta e_x$ is negative, the robots will be commanded to move toward the left wall and halt $\epsilon$ distance from the left wall.
%If $\Delta e_x \ge 0$, the robots will be commanded to move toward the right wall and halt $\epsilon$ distance from the right wall. The epsilon distance prevents the robots from experiencing friction along the vertical wall.
%\item  let $y_{\min} = \min_i r_{iy}$, i.e., $y_{\min}$ be the minimum height of the two robots. Move both robots downward  $y_{\min}$ such that one of the robots touches the bottom wall and hence friction force will prevent this robot from moving left or right.
%\item friction force holds the lower robot in place while the upper robot may move right and left freely to change $\Delta r_x $ to $\Delta e_x$.
%\item  If after the free move of the upper robot $\Delta r_x$ is not $\Delta e_x$, Step (i) will be repeated.  No more than two iterations of step (i) are required.
%\end{enumerate}
%
%Step (ii): Fixing $\Delta r_y$
%Now that $\Delta r_x$ is corrected, $\Delta r_y$ must be corrected:
%\begin{enumerate}
%\item 
%If $\Delta e_y$ is negative, the robots will be commanded to move toward the top wall and halt $\epsilon$ distance from the top wall.
%If $\Delta e_y \ge 0$, the robots will be commanded to move toward the bottom wall and halt $\epsilon$ distance from the bottom wall. The epsilon distance prevents the robots from experiencing friction along the horizontal wall.
%\item  let $x_{\min} = \min_i r_{ix}$, i.e., $x_{\min}$ be the minimum distance of the two robots from the origin along the $x$-axis. Move both robots to the left $x_{\min}$ such that one of the robots  touches the left wall and hence friction force will prevent this robot from moving up or down.
%\item  friction force holds the left robot in place while the right robot may move up or down freely to change $\Delta r_y $  to $\Delta e_y$.
%\item  If after the free move of the right robot $\Delta r_y$ is not $\Delta e_y$, Step (ii) will be repeated.  No more than two iterations of step (ii) are required.
%\end{enumerate}
%Once $\Delta r_x$ and $\Delta r_y$ are set to $\Delta e_x$ and $\Delta e_y$, a straight-line global input moves both robots from $r_1$ and $r_2$ to $e_1$ and $e_2$. 
%

@@

\begin{algorithm}
\caption{GenerateDesired$x$-spacing($s_1,s_2,e_1,e_2,L$)}\label{alg:XControl}
\begin{algorithmic}[1]
\Require knowledge of starting $(s_1,s_2)$ and ending $(e_1,e_2)$ positions of  two robots. 
$(0,0)$ is bottom corner, $s_1$ is topmost robot, 
 $L$ is length of the walls. Current robot positions are $(r_1,r_2)$.
\Ensure   $ r_{1y} - r_{2y}  \equiv s_{1y} - s_{2y} $   %$\Delta y(t) \equiv \Delta y(0)$ 
\State $\epsilon \gets $ small number
\State $ \Delta s_x  \gets s_{1x} - s_{2x} $
\State $ \Delta e_x \gets e_{1x} - e_{2x} $
%\State $ r_1 \gets s_1$, $ r_2 \gets s_2$
\If {$\Delta e_x < 0 $ }
\State move$( L-\epsilon-\max( r_{1x},r_{2x}) ,0)   $ \Comment{To right wall}
\Else 
\State  move$ ( \epsilon-\min( r_{1x},r_{2x}),0 )    $ \Comment{Move to left wall}
\EndIf
\State move$(0, -\min( r_{1y},r_{2y} ))$ \Comment{Move to bottom}
%\State $ r_1 \gets r_1+m$, $ r_2 \gets r_2+m$ \Comment{Apply move}
\If {$\Delta e_x - (r_{1x} - r_{2x} ) > 0 $}
\State move$(\min(|\Delta e_x - \Delta s_x |, L- r_{1x}), 0)$  \Comment{Move right}
\Else
\State move$(-\min(|\Delta e_x - \Delta s_x |, r_{1x}), 0)$\Comment{Move left}
\EndIf 
\State move$(0, \epsilon)$ \Comment{Move up}
%\State $ r_1 \gets r_1+m$, $ r_2 \gets r_2+m$ \Comment{Apply move}
%\State $\Delta r_x = r_{1x} - r_{2x}$
\If {$r_{1x} - r_{2x} \equiv \Delta e_x$} 
%\State   $ m \gets (e_{1x}-r_{1x}, e_{1y}-r_{1y})$
%\State $ r_1 \gets r_1+m$, $ r_2 \gets r_2+m$ \Comment{Apply move}
\State  \Return $(r_1,r_2)$
\Else   
\State \Return GenerateDesired$x$-spacing($r_1,r_2,e_1,e_2,L$)
\EndIf
\end{algorithmic}
\end{algorithm}

%\begin{algorithm}
%\caption{GenerateDesired$y$-spacing($s_1,s_2,e_1,e_2,L$)}\label{alg:YControl}
%\begin{algorithmic}[1]
%\Require Knowledge of starting $(s_1,s_2)$ and ending $(e_1,e_2)$ positions of  two robots. 
%$(0,0)$ is bottom corner, $s_1$ is rightmost robot, 
% $L$ is length of the walls. Current robot positions are $(r_1,r_2)$.
%\Ensure   $ r_{1x} - r_{2x}  \equiv s_{1x} - s_{2x} $   %$\Delta y(t) \equiv \Delta y(0)$ 
%\State $ \Delta s_y  \gets s_{1y} - s_{2y} $
%\State $ \Delta e_y \gets e_{1y} - e_{2y} $
%\State $ r_1 \gets s_1$, $ r_2 \gets s_2$
%\If {$\Delta e_y < 0 $ }
%\State $ m \gets ( L-\max( r_{1y},r_{2y}) ,0)   $ \Comment{Move to top wall}
%\Else 
%\State  $ m \gets ( -\min( r_{1y},r_{2y}),0 )    $ \Comment{Move to bottom wall}
%\EndIf
%\State $m  \gets  m + (0, -\min( r_{1x},r_{2x} ))$ \Comment{Move to left}
%\State $ r_1 \gets r_1+m$, $ r_2 \gets r_2+m$ \Comment{Apply move}
%\If {$\Delta e_y - (r_{1y} - r_{2y} ) > 0 $}
%\State $ m \gets (\min(|\Delta e_y - \Delta s_y |, L- r_{1y}), 0)$  \Comment{Move top}
%\Else
%\State $ m \gets (-\min(|\Delta e_y - \Delta s_y |, r_{1y}), 0)$\Comment{Move bottom}
%\EndIf 
%\State $m  \gets  m + (0, \epsilon)$ \Comment{Move right}
%\State $ r_1 \gets r_1+m$, $ r_2 \gets r_2+m$ \Comment{Apply move}
%\State $\Delta r_y = r_{1y} - r_{2y}$
%\If {$\Delta r_y \equiv \Delta e_y$} 
%\State   $ m \gets (e_{1x}-r_{1x}, e_{1y}-r_{1y})$
%\State $ r_1 \gets r_1+m$, $ r_2 \gets r_2+m$ \Comment{Apply move}
%\State  \Return $(r_1,r_2)$
%\Else   
%\State \Return GenerateDesired$y$-spacing($r_1,r_2,e_1,e_2,L$)
%\EndIf
%\end{algorithmic}
%\end{algorithm}

>>>>

\subsection{Using Boundaries: stable configurations of a swarm}\label{subsec:FluidInTank}
One method to control a swarm's shape in a bounded workspace is to simply push in a given direction until the swarm conforms to the boundary. Like fluid settling in a tank, the stable final configuration minimizes potential energy.
\paragraph{Square workplace}
%The formulas for means $(\bar{x},\bar{y})$, covariance $(\sigma^2_x,\sigma^2_y,\sigma_{xy})$, and correlation $\rho_{xy}$ are as follows: 
We first examine the mean $(\bar{x},\bar{y})$, covariance $(\sigma^2_x,\sigma^2_y,\sigma_{xy})$, and correlation $\rho_{xy}$ of a very large swarm of robots as they move inside a square workplace under the influence of gravity pointing in the direction $\beta$. The swarm is large, but the robots are small in comparison, and together occupy a constant area $A$ ranging from [0,1]. Under a global input such as gravity, they flow like water, moving to a side of the workplace and forming a polygonal shape, as shown in Fig.~\ref{fig:SquareFill},  see also \citep{Zhao2016mathematicaSquare}.

@@

The region of integration $R$ is the polygon containing the swarm. For example, if $A<1/2$ and the force angle is $\beta$, the mean when $R$ is a triangular region in the lower-left corner is:
\begin{align}\label{eq:meanInSquareWorkspaceLL}
\bar{x}(A,\beta) &= \frac{\int_0^{\sqrt{2 A \tan (\beta)}} \left(\int_0^{\cot (\beta) \left(\sqrt{2 A \tan (\beta)}-x\right)} \, dy\right) x \, dx}{A} \nonumber \\
	&=\frac{ \sqrt{2}}{3} \sqrt{A \tan (\beta )}\\
\bar{y}(A,\beta) &= \frac{\int_0^{\sqrt{2 A \tan (\beta)}} \left(\int_0^{\cot (\beta) \left(\sqrt{2 A \tan (\beta)}-x\right)} y \, dy\right) \, dx}{A} \nonumber\\
	&=\frac{\sqrt{2}}{3}  \sqrt{A \cot (\beta )}
\end{align}
The full equations are included in the appendix~\citep{2016arXiv160901830S}, and are summarized in Fig.~\ref{fig:SquareFill}. A few highlights are that the correlation is maximized $\pm$1/2 when the swarm is in a triangular shape. The covariance of a triangle is always $\pm(A/18)$. Variance is minimized in the direction of $\beta$ and maximized orthogonal to $\beta$ when the swarm is in a rectangular shape. The range of mean positions are maximized when $A$ is small.

@@

Let the control input be a vector force $\vec{F}$ with magnitude $F$ and orientation $\theta$ with respect to a line perpendicular to and into the nearest boundary. $N$ is the normal or perpendicular force between the robot and the boundary. The force of friction $F_f$ is nonzero if the robot is in contact with the boundary and  $\sin(\theta) < 0$. The resulting net force on the robot, $F_{\text{\emph{forward}}}$, is aligned with the wall and given by
\begin{align}
F_{\text{\emph{forward}}} &=  F \sin(\theta) - F_f  \nonumber \\
\text{where }  F_f &= \begin{cases}  \mu_f N, &  \mu_f N < F \sin(\theta)  \label{eq:frictionmodel}  \\
F \sin(\theta), & \text{else} \end{cases} \\ %\sign(F \sin(\theta) ) \cdot  \max(0, | F sin\theta |- |F_f|)
\text{and } N &= F \cos(\theta) \nonumber
\end{align}
 Fig.~\ref{fig:friction} shows the resultant forces on two robots when one is touching a wall. 
Though each receives the same inputs,  they experience different net forces.
  For ease of analysis, the following algorithms assume $\mu_f$ is infinite and robots touching the wall are prevented from sliding along the wall.
This means that if one robot is touching the wall and another robot is free, the touching robot will not move when the control input is parallel or into the wall. 
There are many alternate models of friction that also break control symmetry. Fig.~\ref{fig:friction}c shows fluid flow along a boundary.  Fluid in the free-flow region moves uniformly, but flow decreases to zero in the boundary layer.  
\begin{align}
%u(y) = u_0 \left(1- \frac{(y-h)^2}{h^2}\right) = u_0 \frac{y}{h} \left(2- \frac{y}{h}\right) \label{eq:boundarylayerflow} THiS EQUATION IS TERRIBLE! PLOT IT.
F_{forward}(y) &= F - F_f\begin{cases}  \left( \frac{y-h}{h} \right)^2  , &  y<h \label{eq:boundarylayerflow} \\
0, & \text{else} \end{cases}
\end{align}

@@

\section{Algorithms}\label{sec:algorithms}
\subsection{Position Control of $2$ Robots Using Wall Friction}\label{sec:PostionControl2Robots}
\begin{figure*}
\centering
\renewcommand{\figwid}{0.4\columnwidth}
{\begin{overpic}[width =\figwid]{S1_1.png}\put(10,15){$t$ = 0 s}
\end{overpic}
\begin{overpic}[width =\figwid]{S1_2.png}\put(10,15){$t$ = 0.14 s}
\end{overpic}
\begin{overpic}[width =\figwid]{S1_3.png}\put(10,15){$t$  = 0.29 s}
\end{overpic}
\begin{overpic}[width =\figwid]{S1_4.png}\put(10,15){$t$  = 0.43 s}
\end{overpic}
\begin{overpic}[width =\figwid]{S1_5.png}\put(10,15){$t$  = 0.76 s}
\end{overpic}}\\

\caption{\label{fig:shapeControlMathematica1}{Frames from an implementation of Alg.\ \ref{alg:PosControl2Robots}: two robot positioning using walls with infinite friction. The algorithm only requires friction along the bottom and right walls.
Robot initial positions are shown by a crosshair, and final positions by a circled crosshair.  Dashed lines show the shortest route if robots could be controlled independently.  Solid arrows show path given by  Alg.\ \ref{alg:PosControl2Robots}.
Online demonstration and source code at \citep{Shahrokhi2015mathematicaParticle}.
%The bottom row shows an extreme case where the robots must switch position.
}
%\vspace{-2em}
}
\end{figure*}

@@

@@

\begin{algorithm}
\caption{GenerateDesired$x$-spacing($s_1,s_2,e_1,e_2,L$)}\label{alg:XControl}
\begin{algorithmic}[1]
\Require knowledge of starting $(s_1,s_2)$ and ending $(e_1,e_2)$ positions of  two robots. 
$(0,0)$ is bottom corner, $s_1$ is topmost robot, 
 $L$ is length of the walls. Current robot positions are $(r_1,r_2)$.
\State flag $\gets 0$
 \If {$ r_{1y}  \equiv r_{2y}$} 
\State perturb $x$ values
\State flag $\gets 1$
\EndIf
%\Ensure   $ r_{1y} - r_{2y}  \equiv s_{1y} - s_{2y} $   %$\Delta y(t) \equiv \Delta y(0)$ 
\State $\epsilon \gets $ small number
\State $ \Delta s_x  \gets r_{1x} - r_{2x} $
\State $ \Delta e_x \gets e_{1x} - e_{2x} $
\If {$\Delta e_x < 0 $ }
\State move$( L-\epsilon-\max( r_{1x},r_{2x}) ,0)   $ \Comment{Near right wall}
\Else 
\State  move$ ( \epsilon-\min( r_{1x},r_{2x}),0 )    $ \Comment{Near left wall}
\EndIf
\State move$(0, -\min( r_{1y},r_{2y} ))$ \Comment{Move to bottom}
\If {$\Delta e_x - (r_{1x} - r_{2x} ) > 0 $}
\State move$(\min(|\Delta e_x - \Delta s_x |, L- r_{1x}), 0)$  \Comment{Move right}
\Else
\State move$(-\min(|\Delta e_x - \Delta s_x |, r_{1x}), 0)$\Comment{Move left}
\EndIf 
\State move$(0, \epsilon)$ \Comment{Move up}
\If {$r_{1x} - r_{2x} \not\equiv \Delta e_x$} 
\State GenerateDesired$x$-spacing($r_1,r_2,e_1,e_2,L$)
\EndIf

 \If {$ flag == 1$} 
\State unperturb $x$ values
\EndIf

\State  \Return $(r_1,r_2)$
\end{algorithmic}
\end{algorithm}
